# Data visualization with ggplot2

```{r, echo=FALSE, message=FALSE, eval=FALSE}
# Set eval=TRUE to hide all results and figures.
# This sets defaults. Can change this manually in individual chunks.
# Must load knitr so opts_chunk is in search path.
library(knitr)
opts_chunk$set(results="hide", fig.show="hide", fig.keep="none")
```

## Review

Remember an earlier lesson where we introduced basic plotting commands using built-in data and base plotting tools. For example, we made a few plots using Edgar Anderson's famous iris dataset, which measured petal and sepal length and width for several different species of flower.

```{r irisbase}
# Load some data and look at the first few lines
data(iris)
head(iris)

# Make a basic scatter plot
with(iris, plot(Sepal.Length, Petal.Length))
```

When we call `plot` in this way, we are using built-in, or **base** graphics. R's base graphics are powerful and nearly infinitely customizable. 

## ggplot2 

### The Diamonds dataset

Now let's look at a bigger dataset. We're going to be using a data visualization package called **ggplot2** for drawing the plots, and the **ggplot2** package comes with some data we're going to use for this example. 

Recall how to install and load packages. Install the package if you haven't already:

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```

Now let's load the diamonds dataset and take a look at the first few rows and it's structure with commands we learned previously. To learn more about this dataset you can also run `?diamonds`. 

```{r diamondshead}
data(diamonds)
head(diamonds)
str(diamonds)
```

From this we can see this dataset has prices of nearly 54,000 diamonds along with various features about the diamonds, such as the weight, the quality of the cut, the color, the clarity, and measurements of various dimensions. 

### Plotting with ggplot2 using `qplot`

If we wanted to do some exploratory data analysis we might start by plotting the price versus the weight of the diamond.

```{r basediamond}
with(diamonds, plot(carat, price))
```

As we would expect there is definitely a relationship between the size of the diamond and its cost, but how do the other variables (cut, color, clarity) affect the price? We could examine the interrelationships of all these variables using base R graphics, but it could become extremely cumbersome. 

**ggplot2** is a widely used R package that extends R's visualization capabilities. It takes the hassle out of things like creating legends, mapping other variables to scales like color, or faceting plots into small multiples. We'll learn about what all these things mean shortly. To start with, let's produce the same plot as before, but this time using **ggplot2**'s `qplot` function. The `qplot` function is a quick and dirty convenience function to use **ggplot2**, and its syntax will be familiar if you're used to the base `plot` function. We'll cover advanced usage with the `ggplot` function later on.

```{r qplot1}
qplot(carat, price, data = diamonds)
```

The syntax is very similar to R's base graphics where you specify what's on the x and y axes, then give it the name of the data frame you want to use. We see again the strong relationship between the size of the diamond and its price, but the relationship doesn't appear linear. How does the diamond's clarity affect the weight-price relationship? 

#### Faceting and scaling

One option we could use is to color-code the points by their clarity. Here, we pass another `col=` argument with the variable we'd like to use for color-coding:

```{r clarcolor}
qplot(carat, price, data = diamonds, col = clarity)
```

Examining the plot you can see that poor-clarity diamonds (included, small inclusions, etc) fetch a lower price per carat relative to more clear diamonds (very small inclusions, internally flawless, etc). We can see that **ggplot2** color-codes the points using a sensible default color scheme, and automatically draws a legend on the side for us. This requires a good deal of extra error-prone coding using base graphics.

However, with 54,000 points on this plot, there is a good deal of overplotting that obscures how clarity affects the nature of the weight-price relationship. How else might we visualize this data? This is where a *series of small multiples* is helpful. The idea of *small multiples* was popularized by data visualization expert Edward Tufte. The idea is that you create a large grid of small plots, where each plot shows a particular *facet* of the data. Here, each plot in the grid might be price vs. carat for each particular clarity level. You explain to your audience the axes and how to interpret each plot only once, and the audience will immediately understand the rest of the plots.

This can be accomplished easily using **ggplot2**:

```{r facetclar}
qplot(carat, price, data = diamonds, facets = ~ clarity)
```

Here, the `facets` argument expects a forumla object that's constructed with the `~` operator. Here, we've plotted the price vs. weight separately for each level of clarity. We can see what we suspected before. With dirty diamonds (included, and perhaps small inclusions), the weight-price relationship is linear or slightly quadratic. Large diamonds can be purchased rather cheaply. But for very clear diamonds (internally flawless), the relationship is quadratic or even exponential.

Let's examine the weight-price relationship for various color ratings:

```{r facetcol}
qplot(carat, price, data = diamonds, facets = ~ color)
```

Here we see that for whiter diamonds (D, E, F) the price rises more quickly with increasing weight than for yellower diamonds (H, I, J).

We can further facet the plot across two different categorical variables using the same syntax:

```{r facetclarcol, fig.height=10, fig.width=10}
qplot(carat, price, data = diamonds, facets = clarity ~ color)
```

Here we see that the price per carat rises very steeply for very white, very clear diamonds, while the relationship is nearly linear for yellower, more flawed diamonds. We can see that a perfect white diamond averages around $15,000 while a yellow included diamond can be had for only around $2,000.

Finally, we can combind both color-coding and faceting in the same plot. Let's use the same faceting scheme as last time, but color the points by the quality of the diamond's cut.

```{r facetclarcol_colcut, fig.height=10, fig.width=10}
qplot(carat, price, data = diamonds, facets = clarity ~ color, col = cut)
```

This color-coding reveals that clearer, whiter diamonds *generally* have higher quality cuts, but the relationship doesn't appear strong, visually. Looking down the plot toward clearer diamonds you start to see more "Ideal" cuts than at the top, which are the more included diamonds. 

What we've done here in addition to faceting is map a feature of the data (here, the cut quality) onto a scale (here, color). This behavior will work differently depending on whether you're looking at categorical or continuous variables. We can also map features to other *scales* such as `size=`, `shape=`, `linetype=`, or even transparency using `alpha=`. All of these different scales can be combined with each other or with facets, and give you an extremely powerful and easy-to-use graphical toolbox for exploratory data analysis. 

#### Exercise

Now, install the `xx` package, and load the `xx` dataset. Using the techniques we've learned here, use faceting and scaling options to explore how xx, xx, and xx affect the relationship between xx and xx.

***Note:** This is where the *Teaching Software Carpentry* exercise ends, and where the future expansion begins.*

### ggplot2 and the *Grammar of Graphics*

The **ggplot2** package provides an R implementation of Leland Wilkinson's *Grammar of Graphics* (1999). The *Grammar of Graphics* challenges data analysts to think beyond the garden variety plot types (e.g. scatter-plot, barplot) and the consider the components that make up a plot or graphic, such as how data are represented on the plot (as lines, points, etc.), how variables are mapped to coordinates or plotting shape or colour, what transformation or statistical summary is required, and so on. Specifically, **ggplot2** allows users to build a plot layer-by-layer by specifying:

 * a *geom**, which specifies how the data are represented on the plot (points, lines, bars, etc.),
 * *aesthetics* that map variables in the data to axes on the plot or to plotting size, shape, color, etc.,
 * a *stat*, a statistical transformation or summary of the data applied prior to plotting,
 * *facets*, which we've already seen above, that allow the data to be divided into chunks on the basis of other categorical or continuous variables and the same plot drawn for each chunk.

Because **ggplot2** implements a *layered* grammar of graphics, data points and additional information (scatterplot smoothers, confidence bands, etc.) can be added to the plot via additional layers, each of which utilize further geoms, aesthetics, and stats.

To make the best use of **ggplot2** it helps to understand the grammar and how it affects how plots are produced. In addition, it is important to note that **ggplot2** is not a general-purpose plotting tool-kit; you may not be able to achieve certain plots or additions to a figure of they do not map onto concepts included in the layered grammar.

In the examples above, we used **ggplot2**'s convenience function, `qplot`, because it's syntax should be familiar if you've already used base graphics. The `qplot` function did a lot of stuff for us: it created the plot object, added layers, plotted geoms, mapped features to aesthetics, created facets, and displayed the result. From here out, we'll use the `ggplot` function to build everything ourselves. `ggplot` has two required arguments: the *data* used for creating the plot, and an *aesthetic* mapping to describe how variables in said data are mapped to things we can see on the plot.

***More to come.***

## Further **ggplot2** resources

* <http://docs.ggplot2.org/current/>: The official **ggplot2** documentation.
* <http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20>: The **ggplot2** book, by the developer, Hadley Wickham.
* <https://groups.google.com/forum/#!forum/ggplot2>: The **ggplot2** Google Group (mailing list, discussion forum).
* <https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization>: Intermediate Software Carpentry lesson on data visualization with **ggplot2**.
* <http://learnr.wordpress.com/>: A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**.
* <http://stackoverflow.com/questions/tagged/ggplot2>: Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site.
