Software Carpentry UBC Jan 21-22nd 2013. File System Handling from within R.
========================================================

## Part A: Motivation ##

**William S. Noble** _A Quick Guide to Organizing Computaional Biology Projects_ PLoS Comp Bio 2009.

* Set up logical structure for your file system.
* Make work reproducible by avoiding to use the mouse.
* Make directories, files automatic once data set or number of experiments gets large.
* In particular when analysis steps take long, separate analysis from plots and interpretation.

## Part B: warm-up ##

**R** lets us use terminal commands like
```{r warm-up,eval=FALSE}
system("pwd")

system("ls")

system("say Calculations are done! What next?") # Works on Mac. Maybe similar things, or beep, exist on other OS.
```

Problem with those is that command syntax depents on OS. Also **R** can not interpret the results.

```{r why not}
x <- system("pwd") # assignment does not do what we expect.
class(x)
```

Most commands have an equivalent in **R**, so use those instead.

```{r better}
(x <- getwd()) # works
```


## Part C: Directories ## 
```{r directories}
system("ls") # works, but R can not interpret output
help(dir)
y <- dir() # the R way
y[1] 
```

### Task1: get all file names that include the word "file"
```{r task1}
y[grep("file",y)]
#or faster
dir(pattern="file")
```


```{r directory intermediates,eval=FALSE}
dir("../") # folder above working directory (wd is default)
getwd() # but did not change current directory

# more help on dir
apropos("dir")
```

To create new directories use
```{r dir.create,warning=FALSE}
help("dir.create")
dir.create("data")
dir.create("data/2013-01 HIV tests") # "data" needs to be here already
# to detele a directory use
unlink("data/2013-01 HIV tests/",recursive=TRUE) # need TRUE to actually delete directory
dir.create("data/2013-01 HIV tests") # we actually need this directly later, so let's recreate it
```



With this we can easily write a script to set up a logical system of folders that we want to start every new project with.



## PART D: files ##
```{r intro files,eval=FALSE}
apropos("file")
```

**R** lets us create new files and will them with data.
```{r create.files}
file.create("temp.txt")
write.table(runif(10),"temp.txt")
myTable <- read.table("temp.txt") # can specify header=TRUE
print(myTable)
```

### Task 2: Create 10 files and store 1,2,...,10 random numbers ###
```{r task2}
for (j in 1:10){
  currFileName <- paste0("task",j)
  file.create(currFileName)
  write.table(runif(j),currFileName)
}
```


## PART E: moving files ##

Let's remove all the clutter we created.
```{r remove}
file.remove("temp.txt")
```

Instead, we could also move files to other directories, using _file.rename_. A useful command to get the path of the file correctly formated is the function _file.path_, which takes any number of inputs strings and concatenates them to a file path.

```{r move,warning=FALSE}
dir.create("clutter")

# use file.path to create filename
file.path(getwd(),"clutter")

# move one file (and add ".txt"")
file.rename("task1",file.path(getwd(),"clutter","task1.txt"))
```

### task3: use dir(), grep, to move all files with "taskN" to directory clutter ###

```{r task3}
for (filename in dir(pattern="task")){
  file.rename(filename,file.path(getwd(),"clutter",paste0(filename,".txt")))  
}
```


## PART F: work on data ##

```{r prepare Data}
name <- c("A3","B5","A11","C21","C8","33","A4","A18","C7","6","B30")
date <- c("2003-06-11","2011-06-01","2009-12-11","2010-11-03","2003-8-15","2003-2-2","1998-6-20","1997-05-08","1995-10-7","2003-11-15","200-006-09")
type <- c("second gen HIV-1","second gen HIV-2","HIV-2, (second gen)","HIV-1 (second gen)","HIV-1/2, (second gen)","second gen HIV-1/2","second gen HIV-2","HIV-2, (first gen)","HIV-2 (second gen)","first gen HIV-1/2","HIV-1 (first gen)")
test1 <- (runif(11)>0.2)
test2 <- (runif(11)>0.3)
test3 <- (runif(11)>0.4)
test4 <- (runif(11)>0.5)
test5 <- (runif(11)>0.6)
(data <- data.frame(name,type,date,test1,test2,test3,test4,test5))
save(data,file="fileData.RData")
```

### Task 4: move data file to correct folder (2013-01 HIV test). Then, change wd to this folder ###
```{r move Data}
file.rename("fileData.RData",file.path(getwd(),"data","2013-01 HIV tests","fileData.RData"))
setwd(file.path(getwd(),"data","2013-01 HIV tests"))

load("fileData.RData")
```

Now we want to start our analysis:

1. Choose one pattern of HIV tests (e.g. "second" for all second generation tests)
1. For each matching patient calculate the number of positive tests
1. Save this result in folder with name of the _pattern_ and txt file patient name
1. Load the data and create histogram

This sounds a lot like the file _runall_ from the paper mentioned at the intro. Such a file could look like this:

```{r runall,eval=FALSE}
# file runall.R
# patterns we want to match
pattern <- "second"

validIndex <- grep(pattern,data$type)
subData <- data[validIndex,]

dir.create(pattern)

# all the computational work goes here
source("analyzeAndSave.R")
analyzeAndSave(subData,file.path(getwd(),pattern),overWrite=TRUE)

# all the interpretation of results goes here
source("histogramOfPositiveTests")
histogramOfPositiveTests(pattern)
```

Only now we write the function we claim to have above. These could look like
```{r analyzeAndSave}
analyzeAndSave <- function(data,folderName,overWrite=FALSE){
  # saves number of times test was positive in separate files in folderName. Does not overWrite if file is there already (to save computation time)
  n <- nrow(data)

  for (j in 1:n){
    name <- data$name[j]
    fileName <- file.path(folderName,paste0(name,".txt"))
    if (!file.exists(fileName) | overWrite){
        positives <- 0
        for (i in 1:5)
          positives <- positives+data[j,3+i]
        file.create(fileName)
        write.table(positives,fileName)
      }
  }
}
histogramOfPositiveTests <- function(folderName){
# looks for txt files in folderName and creates histogram of number of positive tests
fileNames <- dir(folderName,pattern=".txt$")
firstFile <- TRUE
for (fileName in fileNames){
  currResult  <- read.table(file.path(folderName,fileName))
  if (firstFile){
    res <- currResult
    firstFile <- FALSE
    }
  else
    res <- cbind(res,currResult)
}
hist(c(t(res)))
}
```

And now we can run _runall.R_

```{r see,warning=FALSE}
pattern <- "second"

validIndex <- grep(pattern,data$type)
subData <- data[validIndex,]

dir.create(pattern)

analyzeAndSave(subData,file.path(getwd(),pattern),overWrite=TRUE)
histogramOfPositiveTests(pattern)
```

