Software Carpentry UBC Jan 21-22nd 2013. Regular Expressions.
========================================================

Short script for how to use regular expressions in R.



Regular expressions are used to pre-format data.

```{r prepare Data}
name <- c("A3","B5","A11","C21","C8","33","A4","A18","C7","6","B30")
date <- c("2003-06-11","2011-06-01","2009-12-11","2010-11-03","2003-8-15","2003-2-2","1998-6-20","1997-05-08","1995-10-7","2003-11-15","200-006-09")
type <- c("second gen HIV-1","second gen HIV-2","HIV-2, (second gen)","HIV-1 (second gen)","HIV-1/2, (second gen)","second gen HIV-1/2","second gen HIV-2","HIV-2, (first gen)","HIV-2 (second gen)","first gen HIV-1/2","HIV-1 (first gen)")
(data <- data.frame(name,type,date))
```

Here we are given data from HIV tests. Each row is a patient. There are different kind of tests, for HIV type 1, or 2, or both, taken at different time points.

_The actual test results are missing here as we want to focus on the pre-processing._

**Our task: Make data format consistent.**

Our main tool is the function

```{r grepHelp}
help(grep)
```

Which takes two argument: _pattern_ and _data_ to match pattern against.


## Part A: Dates ##


### Task 1: Extract all tests that were taken in 2003. ###

```{r task1}
grep("2003",data$date)
```

_grep_ returns a list of indices that match the pattern "2003".


### Task 2: Extract all tests that were taken after 2000. ###

We could try "200" as pattern, but then 2010, 2011, etc. would not be captured. If we try "20" we capture too much: this also accepts the day 20th. What we need is "20" followed by two more digits:

```{r task2}
#grep("200",data$date) #-> 201x is not captured

#grep("20",data$date) #-> includes day 20th
grep("20..",data$date) #-> works
```

The "." acts as a wild card that matches any single character.

**Note**
* [0-9] stands for any digit (so does \d). Hence "20[0-9][0-9]" and "20\d\d" would work better (more restrictive) for all tests later than 2000. 
* Sometimes shorter is "20\d{2}", where the _postfix_ {n,m} means "look for previous symbol at least n, at most m times".
* the postfix * mean "zero or more times", the postfix "+" means "one or more times".


As we see, it is easy to miss details and it's cumbersome to compare the output from _grep_ with what we expect.

## regexpal ##

A much better way of checking if your _pattern_ does what you think it does it to go to [this webage](http://regexpal.com/).

**Typical Workflow**

1. Copy subset of data to webpage.

2. Play with pattern until you are confident it matches exactly those entries you want it to.

3. In the above, check that the pattern is not too restrictive, but also not too generous.

4. Once happy with the pattern, get it back into R.


### Task 3: Find all tests done in November (use webpage) ###
```{r task3}
# SOLUTION: -11-

# (not just -11, this also captures 11th day)
```

### Task 4: Find all test done in June (use webpage) ###
```{r task4}
# more tricky, since some are type -6- others are -06-

# SOLUTION: -0{0,1}6-

# shorthand for {0,1} = ?, i.e. -0?6- is equivalent.

# note that -0*6- would accept 6, 06, 006, 0006, etc.
```

### Task 5: Find all test done between February and August (use webpage) ###
```{r task5}
# SOLUTION: -0?[2-8]-
```

## PART B: Names ##

### Task 6: Find all patient names that start with A or C, followed by exactly 2 digits ###
```{r task6}
# SOLUTION: [AC][0-9]{2}
```

The square brackets around "[AC]" indicate that this should be treated as just one element, as opposed to "AC". The omitted operator between the characters in the bracket is *OR*, which you can make specific with "|". Hence "[AC]" "[A|C]", 

Note that "[A-C]"" is any capital letter between A and C. "[a-zA-Z]" would be any single letter.

Now that we've convinced ourselves that our pattern is correct, we can go **back to R.**

```{r AC in R}
grep("[AC][0-9]{2}",data$name)

# return entries, not indices:
data[grep("[AC][0-9]{2}",data$name),]
```

A very similar function that is useful here is _grepl_, which returns a boolean vector.

```{r grepl}
# use grepl instead for TRUE/FALSE
grepl("[AC][0-9]{2}",data$name)
```

This is handy when you want to use logical operations on your results.

### Task 7: Find all patients that match "AC and two digits" and were tested between February and August (use grepl) ###

```{r task7}
(AorCtwoDigits <- grepl("[AC][0-9]{2}",data$name))
(FebToAug <- grepl("-0?[2-8]-",data$date))
(combined <- AorCtwoDigits & FebToAug)
(data[combined,])
```


## Part C: Test Type ##

We can do more than just search for patterns, we can also replace one pattern by another. The key functions here are _gsub_ and _sub_.

```{r gsub}
help(gsub)
gsub("HIV","hiv",data$type)
```

And here is where it gets really exciting. Regular expression can remember what they matchend and use this in the replacement later.

```{r replace}
# The below turns "HIV-1" or "HIV-2" into "Type x hiv" with the correct x!
sub("HIV-([12])","Type \\1 hiv",data$type)
```

The round brackets indicate grouping. They do not match round brackets in the pattern (just as "[]" does not match "[" or "]" directly.) We can refer to the groups by "\\1", "\\2", etc. In order to actually match one of the special characters "(", "]", "?", we need an **escape character**: ``"\("``, "\]", "\?"

_Note that in other languages "\1", "\2" works, but **R** needs two ``"\\"``s._


### Task 8: Improve on the above to shorten each type to HIV-x where x is 1, 2, or ``1/2``.###

```{r task8}
sub(".*HIV-([12](/2)?).*","HIV-\\1",data$type)
```

Note the ".*" to match everything outside the pattern we are looking for. Also note that the second round brackets are used to make "?" interpret "/2" as optional, as compared to "\?2?" which would accept "HIV-1/" and "HIV-12".

### Task 9: Write all types in the format HIV type (i.e. 1,2,or ``1/2``), generation name (i.e. "first"" or "second"). E.g. "``1/2``, second" ###
```{r task9}
# 
newCol <- gsub("(first|second) gen HIV-([12](/2)?)","\\2, \\1",data$type)
newCol <- gsub("HIV-([12](/2)?),? \\((first|second) gen\\)","\\1, \\3",newCol)
(newCol)
```

Note that ``"\\(("`` is used to match an actual left round bracket ``"\\("`` and then to open the thirds group "(".


**Other tricks**
```{r others}
# everything that ends with "gen)", use $
data[grep("gen)$",data$type),]
# everything that starts with 20, use ^
data[grep("^20",data$date),]
```


## Part D: write functions that combine the above ##
```{r regExpHelpers}
cleanType <- function(type){
  #returns Test Type in format "1/2, first"
  
  # first gen HIV-1/2
  cleanType <- gsub("(first|second) gen HIV-([12](/2)?)","\\2, \\1",type)
  
  # HIV-1/2 first gen
  cleanType <- gsub("HIV-([12](/2)?),? \\((first|second) gen\\)","\\1, \\3",cleanType)

  # add more types as data set grows
  
  return(cleanType)
}


extractPatients <- function(cleanPattern,myDataFrame){
  #extracts from myDataFrame those rows whose Type match cleanPattern (e.g. "1, first" or "1/2, second")  
  
  cleanTypes <- cleanType(myDataFrame$type)
  
  extractPatients <- myDataFrame[grepl(cleanPattern,cleanTypes),]
  
  return(extractPatients)
}
```

With this, we can now extract those patients we are interested in in a very short and handy function call:
```{r runHelpers}
extractPatients("1/2, second",data)
```




## Part E: summary ##
1. _grep(pattern,x)_ to match regExp pattern
2. letters, digits match themselves
3. | means or
4. groups are in (group) and refered as ``\\1``, ``\\2``, etc
5. . is any single character
6. postfix * means 0 or more times, + means 1 or more times, ? means 0 or 1 time
7. double backslash as escape character: ``\\(`` to match actual ")"
8. _grepl_ to get logical vector to combine operations
9. _gsub_ to substitute one pattern for another


```

